\section*{Acknowledgments}

This research was performed under appointment of the first author to the
Rickover Fellowship Program in Nuclear Engineering sponsored by Naval Reactor
Division of the U.S. Department of Energy. This work was also supported in part
by the Office of Advanced Scientific Computing Research, Office of Science, US
Department of Energy, under Contract DE-AC02-06CH11357 and by the Consortium for
Advanced Simulation of Light Water Reactors, an Energy Innovation Hub for
Modeling and Simulation of Nuclear Reactors under US Department of Energy
Contract No. DE-AC05-00OR22725.

I would like to express my most sincere gratitude to my advisor, Ben Forget, for
his unconditional support over the years, for granting me pretty much complete
autonomy in pursuing my research, and yet still providing the vision and
necessary guidance to work towards a tangible goal. I could not have asked for a
better advisor, mentor, or friend to have during my years as a graduate student.

I have been most fortunate to also have Kord Smith serve in an advisory capacity
in my work. Kord has helped us tremendously with keeping a focus on realistic
problems and one day bringing OpenMC to a point where it could be used for real
analysis. I am certainly indebted to him, not only for his advice, lectures, and
inspiration, but also for the many drinks that he has subsidized.

It has truly been an honor working closely with Andrew Siegel on the development
of parallel algorithms. His guidance was instrumental in the successful analysis
and implementation of our tally server methodology. I also thank him for his
mentorship on various issues, and look forward to future collaborations
together.

Also deserving acknowledgment are Tom Sutton, Tim Trumbull, and Forrest Brown
for their patience and willingness to answer my many questions on Monte Carlo
methods. My growth as a competant methods developer would not have been possible
without their counsel. I would also like to thank Tim Donovan for his help in
navigating the public utterance process.

Feedback from the nascent OpenMC user and developer community has significantly
improved the usability and performance of the code --- a big thanks to Andrew
Siegel, Bryan Herman, Katie Biegel, Jed Phillips, Steven Fine, Nick Horelik, and
Adam Nelson for all that they've done in this respect.

The five and a half years that I've spent in graduate school have been
incredible thanks to my friends and family all over the country who have
provided constant encouragement in all my pursuits. Finally, I want to thank my
wife and best friend, Audrey, without whose love, understanding, and support,
this would have not been possible. She has pushed me out of my zone of comfort
and opened my eyes to the world, quite literally.
